**Machine learning identifies unique taxa differentiating proximal and distal

### 2017-06-07 submission/manuscript.Rmd
progress with discussion

@@

@@

Discussion of the phyla present and at what levels? Good refs in the gut biogeography paper - mucosa vs lumen enrichment stuff "significant variability between sample locations less than 1 cm apart, which suggests that mucosal microbial populations occur in patches" there is also patchiness in fecal samples themselves: spatial niche partitioning in faeces. . Patchiness/sampling an issue as well as differences in human age, sex, genotype, adn definitely diet, pH, oxygen, etc. (does this go in the intro?)

We identified a range of Simpson diversity values between patients and between each anatomical site. As this measurement takes both number of species and abundance of species into account, it seems that there is a varied level of each component along the human colon. There was no clear pattern of diversity when comparing patients- some patients had a higher level of diversity in the proximal than the distal colon, some patients had the opposite. An early study of microbial diversity and membership of three human subjects found a varied level of diversity in each subject, but that subjects were overall more similar to themselves than to each other (Eckberg 2005). We next used the thetaYC dissimilarity metric to compare the communities present at each anatomical location to each other. We found that while there was a wide range of similarity of each site within a person, the study participants were ultimately more similar to themselves than all of the other participants. This finding is consistent with numerous studies examining the biogeography of the colon in healthy humans (Eckberg 2005, Aguierre de Carcer, 2015, Zhang et al 2013).

To get around the noisiness created by a diverse set of samples, we employed machine learning algorithms to identify microbes specific to each side. For each comparison we identified top X OTUs that were strongly predictive of one site or another. Generally, OTUs identified in each location were consistent with known physiological gradients along the gut axis (cite nature review). For instance, the right mucosa harbored mucosa-associated facultative anaerobes such as _Actinomyces_ and _Enterobacteraceae_ and the aerobic _Psuedomonas_ consistent with the highest oxygen regions of the colon (cite). The left mucosa was far more likely to host strictly anaerobic species such as _Porphyromonas_, _Anaerococcus_, _Finegoldia_ and _Peptoniphilus_. The model was less effective at classifying the right and left lumenal contents, probably because the samples are arguably composed of the same material.

offer another model/schematic?
some of the very high OTUs BLAST to stool residents, like that one bacteriodes (OTU20) - not necessarily anything interesting there.

We detected _F. nucleatum_ and _P. asacharolytica_ in 7 and 5 of our patients, respectively. These bacteria have been shown to be predictive of colorectal cancer in humans (cite Niel) and have oncogenic properties in cell culture and in mice (cite). Interestingly, while _F. nucleatum_ was found on both sides of the colon, _P. asacharolytica_ was only detected in the left mucosa. Not much is known about the distribution of _P. asacharolytica_ but given its documented anaerobic characteristics, it might not be surprising that it resides in the less-oygen-rich left bowel (cite something). In studies examining bacteria on colorectal cancer tumors, _F. nucleatum_ is more commonly detected on right-sided tumors, and distribution of _F. nucleatum_ decreases along the colon to rectum (cite Mima 2016). Of the (8) (40%) individuals positive for _F. nucleatum_ in this study, the bacteria was spread across the right mucosa, left lumen and left mucosa. That said, in a previous study of 1,102 colorectal cancer cases, only 138 (11%) of them contained detectable _F. nucleatum_ at all (Mima 2016). In a separate study of patients with inflammatory bowel disease (IBD), 50% of 22 IBD patients and 17.6% of 34 people without disease contained _F. nucleatum_ in mucosal biopsies (Strauss 2011). These studies have the added caveat of being sampled from prepared or cleaned colons. Thus, the presence of _F. nucleatum_ in a healthy individual is not necessarily linked to the development of future colorectal cancers. Because of the spatial distribution of the _F. nucleatum_ in our sample set, we cannot develop a model for the role of _F. nucleatum_ in the healthy colon. Indeed, rather than the _F. nucleatum_ initiating tumorigenesis, these bacteria may instead arrive "late" to the tumor microenvironment and feed on tumors that have alread begun to develop (Flynn 2016).

what is rate of colonization with these bugs in healthy in literature? how did these bugs factor into our model?

Put this stuff into context with Sears PNAS paper on sidedness

how does bowel prep fit into all of this

Random Forest models have been used by our group and others to increase the detection sensitivity of CRC tumors. Indeed, our group found that a classification model that used microbiome data in combination with Fecal Immunohistochemical Test (FIT) results could correctly identify both carcinoma and adenoma lesions from communities of stool at exit (Baxter et al 2016) and it performed much better than FIT alone. 
- Could try to predict L vs R tumors from exit stool? long term. maybe. probably not. value of getting data about exit stool comparisons. real issue is diagnosing right though. 
- note that we have FIT for all of these people but they're all below limit of detection (except for 2). when we look at models for CRC, cant predict R vs L, so create models for healthy. can also cite niels stuff showing that FIT is representative of stool community. could it also be representative of left? 
- LINK TO EXIT STOOL COMPARISONS.

@@

-what's next?  But really to understand what's going on in the various regions of the gut we need more information than just 16S 
discuss host response papers? what does that mean. link bugs to host response on left and right? future directions or intro

>>>>

add something about sidedness of the gut - embryonic origins, different lymph and blood supplies - more background about male/female differences probably good too

@@

#### Exit stool comparisons model - specifics here or above?

@@

When examining the relative abundance of the different phyla at each site, there is a wide amount of variation for each phyla with communities primarily dominated by the Bacteriodes and Firmicutes. This likely reflects not only the variability between human subjects, casued by differences in age, gender, diet, but also reports of microbial "patchiness" in the gut microbiome. Several previous studies have noted that the bacteria recoverable from the same mucosal sample location can be vastly different when the samples are taken just 1 cm away from each other (cite). Similar patchiness is also observed in lumenal contents and fecal samples themselves; there is observed separation of different interacting microbes along the length of a stool sample, for instance (cite). That said, across our samples the mucosal samples harbor more Proteobacteria, consistent with previous studies comparing mucosal swabs to lumenal content in humans (cite). Thus, the conclusions we can draw from phyla analysis are likely confounded by differences in sampling and patchiness between subjects.

We identified a range of Simpson diversity values between patients and between each anatomical site. As this measurement takes both number of species and abundance of species into account, it seems that there is a varied level of each component along the human colon. There was no clear pattern of diversity when comparing patients- some patients had a higher level of diversity in the proximal than the distal colon, some patients had the opposite. An early study of microbial diversity and membership of three human subjects found a varied level of diversity in each subject, but that subjects were overall more similar to themselves than to each other (Eckberg 2005). We used the thetaYC dissimilarity metric to compare the communities present at each anatomical location to each other. We found that while there was a wide range of similarity of each site within a person, the study participants were ultimately more similar to themselves than all of the other participants. This finding is consistent with numerous studies examining the biogeography of the colon in healthy humans (Eckberg 2005, Aguierre de Carcer, 2015, Zhang et al 2013).

To get around the noisiness created by a diverse set of samples, we built a Random Forest model to identify microbes specific to each side. For each comparison we identified top X OTUs that were strongly predictive of one site or another. Generally, OTUs identified in each location were consistent with known physiological gradients along the gut axis (cite nature review). For instance, the right mucosa harbored mucosa-associated facultative anaerobes such as _Actinomyces_ and _Enterobacteraceae_ and the aerobic _Psuedomonas_ consistent with the highest oxygen regions of the colon (cite). The left mucosa was far more likely to host strictly anaerobic species such as _Porphyromonas_, _Anaerococcus_, _Finegoldia_ and _Peptoniphilus_. The model was less effective at classifying the right and left lumenal contents, probably because the samples are arguably composed of the same material.

We detected _F. nucleatum_ and _P. asacharolytica_ in 7 and 5 of our subjects, respectively. These bacteria have been shown to be predictive of colorectal cancer in humans (cite Niel) and have oncogenic properties in cell culture and in mice (cite). Interestingly, while _F. nucleatum_ was found on both sides of the colon, _P. asacharolytica_ was only detected in the left mucosa. Not much is known about the distribution of _P. asacharolytica_ but given its documented anaerobic characteristics, it might not be surprising that it resides in the less-oygen-rich left bowel (cite something). In studies examining bacteria on colorectal cancer tumors, _F. nucleatum_ is more commonly detected on right-sided tumors, and distribution of _F. nucleatum_ decreases along the colon to rectum (cite Mima 2016). Of the (8) (40%) individuals positive for _F. nucleatum_ in this study, the bacteria was spread across the right mucosa, left lumen and left mucosa. That said, in a previous study of 1,102 colorectal cancer cases, only 138 (11%) of them contained detectable _F. nucleatum_ at all (Mima 2016). In a separate study of patients with inflammatory bowel disease (IBD), 50% of 22 IBD patients and 17.6% of 34 people without disease contained _F. nucleatum_ in mucosal biopsies (Strauss 2011). These studies have the added caveat of being sampled from prepared or cleaned colons. Thus, the presence of _F. nucleatum_ in a healthy individual is not necessarily linked to the development of future colorectal cancers. Because of the spatial distribution of the _F. nucleatum_ in our sample set, we cannot develop a model for the role of _F. nucleatum_ in the healthy colon. Indeed, rather than the _F. nucleatum_ initiating tumorigenesis, these bacteria may instead arrive "late" to the tumor microenvironment and feed on tumors that have alread begun to develop (Flynn 2016). Data examining bacterial biofilms on CRC tumors suggests that _Fusobacteria_ species are more commonly found both on proximal tumors and in biofilms, indicating that it is not only the presence of the bacteria but the organization of the tumor community that contributes to _Fusobacterium's_ role in tumorigenesis.

how does bowel prep fit into all of this- do these match publisehd findings? sort of hard to differentiate bc so many humans?

Microbiome-based diagnostics are increasingly being explored as non-invasive tools to survey for the development of colon cancer. Random Forest models have been used by our group and others to increase the detection sensitivity of CRC tumors. Indeed, our group found that a classification model that used microbiome data in combination with Fecal Immunohistochemical Test (FIT) results could correctly identify both carcinoma and adenoma lesions from communities of stool at exit (Baxter et al 2016) and it performed much better than FIT alone. In this study, all of our subjects were tested using FIT and all but two were found to be below the limit of detection (data not shown). Further work from our lab has shown that microbiome profiling of the FIT cartridge contents sufficiently represented the stool community (cite Niel fit paper). Our work already showed that the stool most accurately matched the samples from the left lumen. So we likely cannot use Random Forest of stool samples to diagnose any changes in the proximal or mucosal communities. But further studies would need to be done. Hard to get a complete set of matched samples from CRC patients though due to discomfort and insurance issues. One caveat of the previous study is that there was not sufficient information to test if a classification model could differentiate between proximal and distal CRC tumors based on exit stool sample alone, but we would hypothesize that would not be effective.

offer another model/schematic?

@@

By revealing specific differences in microbial populations at each location in the gut via sampling an unprepared bowel, we can begin to form hypothesies about how specific host-microbe interactions can affect oncogenesis of proximal and distal CRC tumors. To this point, 16S community profiling studies are a good start but not enough information to probe these questions. Future studies examining metagenomic, metabolomic and host interactions at each site will be useful in further characterizing the host-microbe interactions in the development of proximal and distal colorectal cancer.
